Machine Learning guided workflow for Ribosome Binding Site engineering

### 2021-05-20 paper/main.tex
Updated paper.

One of the main tenets of synthetic biology is design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory. 
Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used \cite{Yeoh2019,Nielsen2016}.
Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements. 
This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.
\mengyan{Maciej:how about ours? same approach or better?}
\maciej{We shouldn't be making comments on ours in the introduction.}
The biggest limitation for the DBTL approach currently is the Learn part of the cycle \mengyan{Maciej: it seems like you are describing TEST?} \maciej{Definitely Learn, Test is only about gathering the data and it has its own limitaions}- there is very limited access to methods and software that can improve and understand designs based on the experimental results.
For example, according to Reeve \emph{et al.} \textcite{Reeve2014}there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
Reported predictions from all of these models are relatively good ($R^2 >0.8$), \mengyan{Maciej: reviewer might wonder why they got high R2 but we don't, e.g. is that because the data? So I recommend either remove this or explain it somewhere.} \maciej{I'll explain it in the discussion}
but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}. 
There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980} . This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016} .

@@

\begin{figure}[!ht]
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.3]{plots/Main_Paper/tsneplot.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
    \end{subfigure}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
    % \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
    \caption{
    \textbf{TIRs of RBS groups examined in this study.} 
    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard, that is TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate. 
    \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
    BPS-NC: base-by-base changes in the non-core region. 
    BPS-C: base-by-base changes in the core region. 
    UNI: Randomly generated sequences with uniform distribution. 
    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
    SD - Shine-Dalgarno sequence.
    Dash line is set to 1 and represents the averaged benchmark sequence TIR for that group. 
    BN - benchmark sequences for all plates. 
    They are not all exactly 1 due to them being shown as separate samples rather than averages.
    \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
    % save the random groups which were shown together due to similar distributions.
    UNI and PPM are merged into Random group and BPS-NC was removed due to changes being made outside the core in this group. 
        \textbf{C)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
    \textbf{D)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
    The area size of the circle represents the experimentally obtained TIR for measured groups.
    \textbf{E)} Exploitation v.s. Exploration for Bandit 1-3. Colors approaching to blue represents exploitation, otherwise represents exploration.}
    \label{fig: Swarmplot and Quantplot}
\end{figure}

@@

Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
% Different randomly generated groups were run together with Bandit 0 (based on literature data) group in Round 0.
% Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
In each round, we measure the TIR of benchmark RBS as internal standard. 
We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).
All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
The Bandit-0 group performed poorly, due to being trained on literature data.
As mentioned before, that data was also exploring the strength of different RBS, however the anatomy of the RBSs in that study was different from ours.
Additionally, the data set was relatively small and covered very small part of the possible experimental space.
However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform beyond Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.
Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms better understanding of the experimental space.
The decreased result in the 0.9th qunatile compared to the max value for Bandit-3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
Figure \ref{fig: Swarmplot and Quantplot}C shows how each Bandit group is over represented in all Bandit results divided into bins. 
\mengyan{Maciej: can you rephrase the last sentence?}
KDE plots have been overlaid to depict the calculated density for each group.
The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the $>0.8$ TIR ratio bins.
\mengyan{do we want to address two-mode pattern (due to the exploration and exploitation.)}
Finally, in \ref{fig: Swarmplot and Quantplot}D we show a tSNE plot depicting the experimental space.
Each RBS is located on the plot according to its distance to other RBSs as calculated by our kernel function.
We can see the RBSs recommended by Bandits groups have covered majority of the design space. 
A number of clusters were targeted by our recommendation algorithm.
For example, the cluster positioned in the top centre labelled as "G rich" has been actively recommended by the algorithm.
More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.
\mengyan{explain the exploration and exploitation balance?}

@@

\mengyan{Maciej: we can add one paragraph and figures here to show what the highlight/benefits of our library, and potential help for future research, if any. Comparison with existing library, if any.} \maciej{We can't produce a good library figure, but have added a paragraph below.}

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[scale=0.75]{plots/Main_Paper/SD.pdf}
    \caption{\textbf{Distribution of standard deviation of tested samples} Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation). TIR is given in terms of raw number.}
    \label{fig: SD}
\end{figure}

\mengyan{Maciej: I realised we haven't explain what is a biological replicates in our study, and why we need that (ML view, I can add this). If providing biological replicates is not common in existing library, then we can emphasis that as a contribution/characteristics of our library.}
\maciej{Biological replicate is again a standard term for biology, I don't think we need to explain what that is. I have added why that is important for our use case though.}

>>>>

One of the main tenets of synthetic biology is design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory. 
Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used \cite{Yeoh2019,Nielsen2016}.
Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements. 
This means that many designers have to rely on known and characterised parts that may not be optimal for their constructs.
The problem with this approach is that such part libraries can also be unreliable due to poor reliability of methods used to obtain them.

The biggest limitation for the DBTL approach currently is the Learn part of the cycle - there is very limited access to methods and software that can improve and understand designs based on the experimental results.
For example, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
Reported predictions from all of these models are relatively good ($R^2 >0.8$), 
% \mengyan{Maciej: reviewer might wonder why they got high R2 but we don't, e.g. is that because the data? So I recommend either remove this or explain it somewhere.} \maciej{I'll explain it in the discussion}
but they come with a number of caveats: i) they rely on calculations of free energies that can be hard to estimate with high precision ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomenons taken into account, but this can lead to paradoxically decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016} and iii) by using deterministic coefficients to calculate energies one disregards often stochastic nature of processes in the cells which again increases perceived prediction error \cite{Goss1998}. 
There are also sources showing that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980} . This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016} .

@@

\begin{figure}[!ht]
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.3]{plots/Main_Paper/tsneplot.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
    \end{subfigure}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
    % \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
    \caption{
    \textbf{TIRs of RBS groups examined in this study.} 
    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard, that is TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate. 
    \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
    BPS-NC: base-by-base changes in the non-core region. 
    BPS-C: base-by-base changes in the core region. 
    UNI: Randomly generated sequences with uniform distribution. 
    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
    SD - Shine-Dalgarno sequence.
    Dash line is set to 1 and represents the averaged benchmark sequence TIR for that group. 
    BN - benchmark sequences for all plates. 
    They are not all exactly 1 due to them being shown as separate samples rather than averages.
    \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
    % save the random groups which were shown together due to similar distributions.
    UNI and PPM are merged into Random group and BPS-NC was removed due to changes being made outside the core in this group. 
        \textbf{C)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
    \textbf{D)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
    The area size of the circle represents the experimentally obtained TIR for measured groups.
    \textbf{E)} Exploitation v.s. Exploration for Bandit 1-3. Blue-hued points represent exploitation, those hued red represent exploration.}
    \label{fig: Swarmplot and Quantplot}
\end{figure}

@@

Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
% Different randomly generated groups were run together with Bandit 0 (based on literature data) group in Round 0.
% Subsequently, the following rounds from 1 to 3 consisted of only Bandit recommended designs.
In each round, we measure the TIR of benchmark RBS as internal standard. 
We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).
All Round 0 groups (BPS-NC, BPS-C, UNI, PPM, Bandit-0) have performed worse than our benchmark sequence in terms of TIR. 
The Bandit-0 group performed poorly, due to being trained on literature data.
As mentioned before, that data was also exploring the strength of different RBS, however the anatomy of the RBSs in that study was different from ours.
Additionally, the data set was relatively small and covered very small part of the possible experimental space.
However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform beyond Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.
Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms better understanding of the experimental space.
The decreased result in the 0.9th qunatile compared to the max value for Bandit-3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
Figure \ref{fig: Swarmplot and Quantplot}C shows how each Bandit group is over represented in all Bandit results divided into bins. 
\mengyan{Maciej: can you rephrase the last sentence?}
KDE plots have been overlaid to depict the calculated density for each group.
The increase of prevalence of later bandit groups in the higher bins is evident, especially for Bandit 2 and 3 constituting the bulk of results in the $>0.8$ TIR ratio bins.
\mengyan{do we want to address two-mode pattern (due to the exploration and exploitation.)}
In \ref{fig: Swarmplot and Quantplot}D we show a tSNE plot depicting the experimental space.
Each RBS is located on the plot according to its distance to other RBSs as calculated by our kernel function.
We can see the RBSs recommended by Bandits groups have covered majority of the design space. 
A number of clusters were targeted by our recommendation algorithm.
For example, the circled clusters 
% positioned in the top centre
labelled as "G-Rich Clusters" have been actively recommended by the algorithm.
More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.
Finally, in Figure \ref{fig: Swarmplot and Quantplot} E) we coloured the swarmplot for Bandit 1-3 groups as the scores representation the exploration-exploitation level. The scores are calculated based on the projection of prediction standard deviation and mean pairs (for details see Supplementary). 
We can see the RBSs with high TIRs (blue, with high predicted mean) tend to exploit the design space; the explorative points (red, with high predicted standard deviation) gives relatively low TIR but expands the knowledge of about the unknown design space. 
Another observation is that a higher ratio of the points in Bandit-3 compared the other two groups are exploitative. 
This is an expected behaviour of our algorithm since we have set a large exploitation rate for the last group (Bandit-3).

@@

% REVIEW:
% \mengyan{Maciej: we can add one paragraph and figures here to show what the highlight/benefits of our library, and potential help for future research, if any. Comparison with existing library, if any.} \maciej{We can't produce a good library figure, but have added a paragraph below.}

@@

\begin{figure}
     \centering
     \begin{subfigure}[b]{0.49\textwidth}
         \centering
         \caption{}
         \includegraphics[scale=0.4]{plots/Main_Paper/SD.pdf}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.49\textwidth}
         \centering
         \caption{}
         \includegraphics[scale=0.6]{paper/plots/Main_Paper/SDvsTIR.pdf}
     \end{subfigure}
     \caption{\textbf{Distribution of standard deviation of tested samples} A) Joint plot showing the distribution of relative error calculated as the average TIR from 6 replicates divided by the given samples standard deviation (coefficient of variation). B) Scatter plot showing the standard deviation for each RBS plotted against its averaged TIR. Standard deviation and average TIR is given in terms of raw number.}
     \label{fig:Library characteristics}
\end{figure}

% REVIEW:
% \mengyan{Maciej: I realised we haven't explain what is a biological replicates in our study, and why we need that (ML view, I can add this). If providing biological replicates is not common in existing library, then we can emphasis that as a contribution/characteristics of our library.}
% \maciej{Biological replicate is again a standard term for biology, I don't think we need to explain what that is. I have added why that is important for our use case though.}
